---
title: "Hypothesis 1, 2, and 3"
author: "Tanner M. Barnes"
date: "`r Sys.Date()`"
output: html_document
---

Get data ready for modeling.

## Load Libraries

```{r setup, include = FALSE}
# This chunk will load necessary packages and data
library(tidyverse)
library(readxl)
library(brms)
library(rstan)
library(tidyr)
library(purrr)
library(broom)

Sys.setenv(PATH = paste("C:/Users/Tanner/OneDrive - Michigan Technological University/PhD/rtools44/x86_64-w64-mingw32.static.posix/bin",
                        "C:/Users/Tanner/OneDrive - Michigan Technological University/PhD/rtools44/usr/bin", 
                        Sys.getenv("PATH"), 
                        sep = ";"))
```

Step 1: Load the data and nest it to create the recovery estimate per site. 

Step 2: Fit a linear regression model for each site

Step 3: Tidy the model outputs

Step 4: Select the slope and intercept for each site and combine them properly

Step 5: Fix the NA values by filling them appropriately

Step 6: Add the slope to the original data frame 
```{r load-data, include = TRUE}

df <- read_excel("C:/Users/Tanner/OneDrive - Michigan Technological University/PhD/Chapter1/actual_data.xlsx")

nest <- df %>% filter(year >= decrease_year) %>% select(site, normalized_count, relative_year) %>% group_by(site) %>% nest()

nested_data <- nest %>%
  mutate(model = map(data, ~ {
    df <- .x
    # Ensure 'year' and 'count' are numeric
    df <- df %>%
      mutate(year = as.numeric(relative_year),
             count = as.numeric(normalized_count))
    # Fit the linear model
    lm(count ~ year, data = df)
  }))

# Step 3: Tidy the model outputs
tidied_data <- nested_data %>%
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied)

# Step 4: Select the slope and intercept for each site and combine them properly
regression_results <- tidied_data %>%
  filter(term %in% c("year", "(Intercept)")) %>%  # Keep both slope and intercept terms
  pivot_wider(names_from = term, values_from = estimate) %>%  # Spread terms into columns
  select(site, intercept = `(Intercept)`, slope = year)  # Select and rename the columns

# Fix NA values by filling them appropriately
regression_results <- regression_results %>%
  group_by(site) %>%
  summarize(
    intercept = max(intercept, na.rm = TRUE),
    slope = max(slope, na.rm = TRUE)
  ) %>%
  ungroup()

regression_results <- regression_results %>% 
mutate(slope = ifelse(is.na(slope), 0, slope),
intercept = ifelse(is.na(intercept), 0, intercept))

model_data <- df %>%
  left_join(regression_results, by = "site") %>% 
  mutate(crash = 1 - (min_count / mean_count), 
  log_passage = log(passage_length)) %>% 
  select(site, slope, intercept, crash, mean_temp, decrease_year, recovery_years, min_count, 
  last_count, mean_count, last_year, passage_length, log_passage) %>%
  filter(site != "Tippy Dam") %>%
  group_by(site) %>% slice(1)
```


# Hypothesis 1
Show that recovery rate and rate of crash are related

```{r frequentist-linear-model-1, include = TRUE}
m1 <- lm(slope ~ crash, model_data)
summary(m1)
```

```{r Visulization-1, include = TRUE}
ggplot(model_data, aes(x=crash, y=slope)) + 
geom_point(aes(fill = mean_temp, size = last_count), shape = 21, color = "black", stroke = 0.5) + 
geom_smooth(method = "lm", se = FALSE, color = "black") +
scale_fill_gradientn(colors = colors, values = scales::rescale(c(0, 0.5, 1)), name = "Mean\nTemperature") +
scale_size_continuous(name = "Last Survey\nPopulation Count") + 
labs(title = "The recovery rate and the population crash are related",
x = "Population Crash (1 - (minimum / mean before WNS))",
y = "Slope (Recovery Rate)") + 
theme_bw() +
theme(
  plot.title = element_text(size = 10, face = "bold", hjust = 0.5, vjust = 1, lineheight = 0.8)) +
  annotate("text", x = Inf, y = Inf, label = "Adjusted R² = 0.3667", 
           hjust = 2.5, vjust = 2, size = 3, color = "black", fontface = "italic")
```

# Hypothesis 2

### Bayesian Modeling 
```{r Hypothesis-2-models, include = FALSE}

# Only include sites that are recovering
model_data_recover <- model_data %>% filter(slope > 0)

# Define the prior
prior <- c(
  prior(normal(0, 1), class = "b", coef = "mean_temp"),
  prior(normal(0,1), class = "Intercept"))

prior1 <- prior(normal(0,1), class = "Intercept")

slope_model <- brm(
  formula = slope ~ mean_temp,
  data = model_data_recover,
  family = Gamma(link = "log"),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)
```

```{r slope_model, include = TRUE}
summary(slope_model)
bayesian_r2 <- bayes_R2(slope_model)
print(bayesian_r2)
```

```{r Hypothesis-2-visulization, include = TRUE}
colors <- c("blue", "white", "red")
b_r2_slope <- bayesian_r2[1, "Estimate"]
bayesian_r2_ci <- quantile(bayesian_r2, probs = c(0.025, 0.975))

model_data_ungroup <- model_data_recover %>% ungroup()

# Create a grid of mean_temp
prediction_grid_s <- model_data_ungroup %>%
  select(mean_temp, recovery_years) %>%  # Ensure only relevant columns are selected
  tidyr::expand(mean_temp = seq(min(mean_temp, na.rm = TRUE), max(mean_temp, na.rm = TRUE), length.out = 100)) %>%
  mutate(recovery_years = median(model_data_ungroup$recovery_years, na.rm = TRUE))

# Generate predictions using the fitted values from the model
predicted_slope_values <- fitted(slope_model, newdata = prediction_grid_s, summary = TRUE)[, "Estimate"]

# Add the predicted values to the dataset
prediction_grid_s <- prediction_grid_s %>%
  mutate(predicted_slope = predicted_slope_values)

  # Plot the slope model
ggplot(model_data_ungroup, aes(x = mean_temp, y = slope)) +
  geom_point(alpha = 0.5) +  # Plot observed data
  geom_line(data=prediction_grid_s, aes(mean_temp, y = predicted_slope), color = "darkblue", linewidth = 1) +
  scale_size_continuous(name = "Last Survey\nPopulation Count") + 
  labs(title = "Mines with colder mean temperatures have higher recovery rates for Myotis bats",
       x = "Mean Temperature (C)",
       y = "Recovery Rate (Slope)") +
  annotate("text", x = Inf, y = Inf, label = paste("Bayesian R² =", round(b_r2_slope, 4)), 
           hjust = 2.75, vjust = 1.5, size = 4, color = "black") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12),        # Title font size
    axis.title.x = element_text(size = 10),      # X-axis title font size
    axis.title.y = element_text(size = 10),      # Y-axis title font size
    axis.text.x = element_text(size = 8),        # X-axis text font size
    axis.text.y = element_text(size = 8),        # Y-axis text font size
    legend.title = element_text(size = 10),      # Legend title font size
    legend.text = element_text(size = 8),
    panel.grid = element_blank()
  )
```
