---
title: "Supplemental Information"
author: "Tanner M. Barnes"
date: "`r Sys.Date()`"
output: html_document
---

### Load the data

```{r mines_used, include = FALSE}
library(readxl)
library(tidyverse)
library(broom)
library(tidyr)
library(purrr)
df <- read_excel("C:/Users/Tanner/OneDrive - Michigan Technological University/PhD/Chapter1/actual_data.xlsx")

nest <- df %>% filter(year >= decrease_year) %>% select(site, count, normalized_count, relative_year, year_from_min) %>% group_by(site) %>% nest()

# STEP 2: Fit a linear regression model for each site
nested_data <- nest %>%
  mutate(model = map(data, ~ {
    df <- .x
    # Ensure 'year' and 'count' are numeric
    df <- df %>%
      mutate(year = as.numeric(year_from_min), 
              count = as.numeric(count))
    # Fit the linear model
    lm(count ~ year, data = df)
  }))

# Step 3: Tidy the model outputs
tidied_data <- nested_data %>%
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied)

# Step 4: Select the slope and intercept for each site and combine them properly
regression_results <- tidied_data %>%
  filter(term %in% c("year", "(Intercept)")) %>%  # Keep both slope and intercept terms
  pivot_wider(names_from = term, values_from = estimate) %>%  # Spread terms into columns
  select(site, intercept_count = `(Intercept)`, slope_count = year)  # Select and rename the columns

# Fix NA values by filling them appropriately
regression_results <- regression_results %>%
  group_by(site) %>%
  summarize(
    intercept_count = max(intercept_count, na.rm = TRUE),
    slope_count = max(slope_count, na.rm = TRUE)
  ) %>%
  ungroup()

regression_results <- regression_results %>% 
mutate(slope_count = ifelse(is.na(slope_count), 0, slope_count),
intercept_count = ifelse(is.na(intercept_count), 0, intercept_count))

# Step 6: Add the slope to the original data frame
final_data <- df %>%
  left_join(regression_results, by = "site")

# Create the trend line data using slope and intercept
your_data <- final_data %>%
  mutate(predicted_count = intercept_count + slope_count * relative_year)
```

### Visualize the mines used and their recovery slopes

```{r visualize_mines_used, include = TRUE}
colors <- c("blue", "white", "red")
your_data %>%
group_by(site) %>% 
mutate(year = as.numeric(year)) %>% 
filter(!is.na(count)) %>% 
  ggplot(aes(x = year, y = count, group = site)) +
  geom_line(show.legend = FALSE) +
  geom_line(aes(y=predicted_count, color = mean_temp), size = 1.2, show.legend = FALSE) +
  scale_color_gradientn(colors = colors, values = scales::rescale(c(0, 0.5, 1)), name = "Mean\nTemperature") +
  facet_wrap(~site, scales = "free_y") +  # Facet by site with free y-axis scales
  scale_x_continuous(
    breaks = seq(min(your_data$year), max(your_data$year), by = 4),
    labels = function(x) sprintf("%02d", x %% 100)
  ) +
  geom_vline(xintercept = 2014, color = "red", linetype = "dashed", size = 0.5) +  # Adjust size here
  theme_dark() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Recovering mines with recovery estimate overlayed the normalized count",
       x = "Year",
       y = "Normalized Population Count")
```

## Model selection

```{r prepare_data, include = FALSE}
####### FIGURE WHICH MODELS ARE BEST? ###############
library(tidyverse)
library(readxl)
library(brms)
library(rstan)
library(tidyr)
library(purrr)
library(broom)
df <- read_excel("C:/Users/Tanner/OneDrive - Michigan Technological University/PhD/Chapter1/actual_data.xlsx")

nest <- df %>% filter(winter_year >= decrease_year) %>% select(site, normalized_count, relative_year) %>% group_by(site) %>% nest()

nested_data <- nest %>%
  mutate(model = map(data, ~ {
    df <- .x
    # Ensure 'year' and 'count' are numeric
    df <- df %>%
      mutate(year = as.numeric(relative_year),
             count = as.numeric(normalized_count))
    # Fit the linear model
    lm(count ~ year, data = df)
  }))

# Step 3: Tidy the model outputs
tidied_data <- nested_data %>%
  mutate(tidied = map(model, tidy)) %>%
  unnest(tidied)

# Step 4: Select the slope and intercept for each site and combine them properly
regression_results <- tidied_data %>%
  filter(term %in% c("year", "(Intercept)")) %>%  # Keep both slope and intercept terms
  pivot_wider(names_from = term, values_from = estimate) %>%  # Spread terms into columns
  select(site, intercept = `(Intercept)`, slope = year)  # Select and rename the columns

# Fix NA values by filling them appropriately
regression_results <- regression_results %>%
  group_by(site) %>%
  summarize(
    intercept = max(intercept, na.rm = TRUE),
    slope = max(slope, na.rm = TRUE)
  ) %>%
  ungroup()

regression_results <- regression_results %>% 
mutate(slope = ifelse(is.na(slope), 0, slope),
intercept = ifelse(is.na(intercept), 0, intercept))

model_data <- df %>%
  left_join(regression_results, by = "site") %>% 
  mutate(crash = 1 - (min_count / mean_count), 
  log_passage = log(passage_length)) %>% 
  select(site, slope, intercept, crash, mean_temp, decrease_year, recovery_years, min_count, 
  last_count, mean_count, last_year, passage_length, log_passage) %>%
  filter(site != "Tippy Dam") %>%
  group_by(site) %>% slice(1)

model_data$slope_weighted <- model_data$slope * sqrt(model_data$last_count)

model_data_recover <- model_data %>% filter(slope > 0)
```

## Crash Model Selection

```{r crash_models, include = FALSE}
null_model_crash <- brm(
  formula = crash ~ 1,
  data = model_data_recover,
  family = Beta(),
  prior = prior1,
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99))

crash_model <- brm(
  formula = crash ~ mean_temp + I(mean_temp^2),
  data = model_data_recover,
  family = Beta(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99))

  crash_model_linear <- brm(
  formula = crash ~ mean_temp,
  data = model_data_recover,
  family = Beta(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99))

  crash_model1 <- brm(
  formula = crash ~ mean_temp + log_passage,
  data = model_data_recover,
  family = Beta(),
  prior = prior,
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99))
```

### Compare crash models
```{r compare_crash_models}
loo_crash_null <- loo(null_model_crash, moment_match = TRUE)
loo_crash_model <- loo(crash_model, moment_match = TRUE)
loo_crash_model_linear <- loo(crash_model_linear, moment_match = TRUE)
loo_crash_model1 <- loo(crash_model1, moment_match = TRUE)

# Extract ELPD values from each model
elpd_null_crash <- loo_crash_null$estimates["elpd_loo", "Estimate"]
elpd_crash_model <- loo_crash_model$estimates["elpd_loo", "Estimate"]
elpd_crash_model_linear <- loo_crash_model_linear$estimates["elpd_loo", "Estimate"]
elpd_crash_model1 <- loo_crash_model1$estimates["elpd_loo", "Estimate"]

# Get the comparison results (elpd_diff and se_diff)
loo_comparison1 <- loo_compare(loo_crash_null, loo_crash_model, loo_crash_model_linear, loo_crash_model1)

# Create a summary table with ELPD, elpd_diff, and se_diff
comparison_table1 <- data.frame(
  ELPD = c(elpd_null_crash, elpd_crash_model, elpd_crash_model_linear, elpd_crash_model1),
  elpd_diff = loo_comparison1[, "elpd_diff"],
  se_diff = loo_comparison1[, "se_diff"]
)

# Print the table
print(comparison_table1)
```

### Posterior Probability Check and Fitted vs Residual plots to check model assumptions for Crash models

```{r crash_model_residual_plots, include = TRUE}
pp_check(crash_model_linear, type = "dens_overlay") + ggtitle("Posterior Check for the crash model (crash ~ mean temperature)")

residuals <- residuals(crash_model_linear)
fitted_values <- fitted(crash_model_linear)
residuals_data <- data.frame(
  Fitted = fitted_values[, "Estimate"],  # Fitted estimates
  Residuals = residuals[, "Estimate"])    # Residual estimates

# Residuals vs Fitted plot
ggplot(residuals_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

  # Q-Q plot of residuals
ggplot(residuals_data, aes(sample = Residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()
```

## Slope model selection

```{r model_selection, include = FALSE}
slope_model <- brm(
  formula = slope ~ mean_temp + offset(recovery_years),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

slope_model1 <- brm(
  formula = slope ~ mean_temp + I(mean_temp^2) + offset(recovery_years),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

slope_model2 <- brm(
  formula = slope ~ mean_temp + log_passage + offset(recovery_years),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

no_offset <- brm(
  formula = slope ~ mean_temp,
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

no_offset1 <- brm(
  formula = slope ~ mean_temp + log_passage,
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

no_offset2 <- brm(
  formula = slope ~ mean_temp + I(mean_temp^2),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99))
```

### Slope model summaries

``` {r model_summaries, include = TRUE}
summary(slope_model)

summary(slope_model1)

summary(slope_model2)

summary(no_offset)

summary(no_offset1)

summary(no_offset2)
```

### Model comparison

```{r model_comparison, include = TRUE}
############################################################################################
##################### Model Comparison ##################################################
########################################################################################
# LOO-CV tends to be more stable, especially in small datasets or models with influencial observations
loo_slope_model <- loo(slope_model, moment_match = TRUE)
loo_slope_model1 <- loo(slope_model1, moment_match = TRUE)
loo_slope_model2 <- loo(slope_model2, moment_match = TRUE)
loo_no_offset <- loo(no_offset, moment_match = TRUE)
loo_no_offset1 <- loo(no_offset1, moment_match = TRUE)
loo_no_offset2 <- loo(no_offset2, moment_match = TRUE)

# Extract ELPD values from each model
elpd_slope_model <- loo_slope_model$estimates["elpd_loo", "Estimate"]
elpd_slope_model1 <- loo_slope_model1$estimates["elpd_loo", "Estimate"]
elpd_slope_model2 <- loo_slope_model2$estimates["elpd_loo", "Estimate"]
elpd_no_offset <- loo_no_offset$estimates["elpd_loo", "Estimate"]
elpd_no_offset1 <- loo_no_offset1$estimates["elpd_loo", "Estimate"]
elpd_no_offset2 <- loo_no_offset2$estimates["elpd_loo", "Estimate"]

loo_comparison <- loo_compare(loo_slope_model, loo_slope_model1, loo_slope_model2,
loo_no_offset, loo_no_offset1, loo_no_offset2)
# Create a summary table with ELPD, elpd_diff, and se_diff
comparison_table <- data.frame(
  ELPD = c(elpd_slope_model, elpd_slope_model1, elpd_slope_model2, 
  elpd_no_offset, elpd_no_offset1, elpd_no_offset2),
  elpd_diff = loo_comparison[, "elpd_diff"],
  se_diff = loo_comparison[, "se_diff"]
)
# Print the table
print(comparison_table)
```

### Check data families with posterior probability check

```{r check_families, include = FALSE}
############ USE PP_CHECK() to check each of the families
# Fit models with different families
model_gaussian <- brm(
  formula = slope ~ mean_temp,
  data = model_data_recover,
  family = gaussian(),
  chains = 4, iter = 4000, warmup = 1000, control = list(adapt_delta = 0.99))

model_lognormal <- brm(
  formula = slope ~ mean_temp,
  data = model_data_recover,
  family = lognormal(),
  chains = 4, iter = 4000, warmup = 1000, control = list(adapt_delta = 0.99))

model_gamma <- brm(
  formula = slope ~ mean_temp,
  data = model_data_recover,
  family = Gamma(link = "log"),
  chains = 4, iter = 4000, warmup = 1000, control = list(adapt_delta = 0.99))

model_student <- brm(
  formula = slope ~ mean_temp,
  data = model_data_recover,
  family = student(),
  chains = 4, iter = 4000, warmup = 1000, control = list(adapt_delta = 0.99))
```


```{r slope_model_family_likelihoods, include = TRUE}
# Posterior predictive check for Gaussian model
pp_check(model_gaussian, type = "dens_overlay") + ggtitle("Gaussian Model")
# Posterior predictive check for Log-Normal model
pp_check(model_lognormal, type = "dens_overlay") + ggtitle("Log-Normal Model")
# Posterior predictive check for Gamma model
pp_check(model_gamma, type = "dens_overlay") + ggtitle("Gamma Model")
# Posterior predictive check for student model
pp_check(model_student, type = "dens_overlay") + ggtitle("Student Model")
```

### Check model assumptions for Slope models

```{r model_assumptions, include = TRUE}
# Step 1: Extract Residuals and fitted values
# Extract residuals
residuals <- residuals(slope_model)
# Extract fitted values
fitted_values <- fitted(slope_model)

# Create a data frame with residuals and fitted values
residuals_data <- data.frame(
  Fitted = fitted_values[, "Estimate"],  # Fitted estimates
  Residuals = residuals[, "Estimate"]    # Residual estimates
)

# Residuals vs Fitted plot
ggplot(residuals_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# Histogram of residuals
ggplot(residuals_data, aes(x = Residuals)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Frequency") +
  theme_minimal()

# Q-Q plot of residuals
ggplot(residuals_data, aes(sample = Residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()
```


### Test weighted slopes (slope * sqrt(last survey count))

```{r fit_weighted_models, include = FALSE}
slope_weighted_model <- brm(
  formula = slope_weighted ~ mean_temp + offset(recovery_years),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

slope_weighted_model1 <- brm(
  formula = slope_weighted ~ mean_temp + I(mean_temp^2) + offset(recovery_years),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

slope_weighted_model2 <- brm(
  formula = slope_weighted ~ mean_temp + log_passage + offset(recovery_years),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

no_weighted_offset <- brm(
  formula = slope_weighted ~ mean_temp,
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

no_weighted_offset1 <- brm(
  formula = slope_weighted ~ mean_temp + log_passage,
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99)
)

no_weighted_offset2 <- brm(
  formula = slope_weighted ~ mean_temp + I(mean_temp^2),
  data = model_data_recover,
  family = student(),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  control = list(adapt_delta = 0.99))
```

### Weighted slope model summaries

``` {r weighted_model_summaries, include = TRUE}
summary(slope_weighted_model)

summary(slope_weighted_model1)

summary(slope_weighted_model2)

summary(no_weighted_offset)

summary(no_weighted_offset1)

summary(no_weighted_offset2)
```

### Weighted slope model comparisons

```{r weighted_model_comparison, include = TRUE}
############################################################################################
##################### Model Comparison ##################################################
########################################################################################
# LOO-CV tends to be more stable, especially in small datasets or models with influencial observations
loo_slope_weighted_model <- loo(slope_weighted_model, moment_match = TRUE)
loo_slope_weighted_model1 <- loo(slope_weighted_model1, moment_match = TRUE)
loo_slope_weighted_model2 <- loo(slope_weighted_model2, moment_match = TRUE)
loo_no_weighted_offset <- loo(no_weighted_offset, moment_match = TRUE)
loo_no_weighted_offset1 <- loo(no_weighted_offset1, moment_match = TRUE)
loo_no_weighted_offset2 <- loo(no_weighted_offset2, moment_match = TRUE)

# Extract ELPD values from each model
elpd_slope_weighted_model <- loo_slope_weighted_model$estimates["elpd_loo", "Estimate"]
elpd_slope_weighted_model1 <- loo_slope_weighted_model1$estimates["elpd_loo", "Estimate"]
elpd_slope_weighted_model2 <- loo_slope_weighted_model2$estimates["elpd_loo", "Estimate"]
elpd_no_weighted_offset <- loo_no_weighted_offset$estimates["elpd_loo", "Estimate"]
elpd_no_weighted_offset1 <- loo_no_weighted_offset1$estimates["elpd_loo", "Estimate"]
elpd_no_weighted_offset2 <- loo_no_weighted_offset2$estimates["elpd_loo", "Estimate"]

loo_comparison <- loo_compare(loo_slope_weighted_model, loo_slope_weighted_model1, loo_slope_weighted_model2,
loo_no_weighted_offset, loo_no_weighted_offset1, loo_no_weighted_offset2)
# Create a summary table with ELPD, elpd_diff, and se_diff
comparison_table2 <- data.frame(
  ELPD = c(elpd_slope_weighted_model, elpd_slope_weighted_model1, elpd_slope_weighted_model2, 
  elpd_no_weighted_offset, elpd_no_weighted_offset1, elpd_no_weighted_offset2),
  elpd_diff = loo_comparison[, "elpd_diff"],
  se_diff = loo_comparison[, "se_diff"]
)
# Print the table
print(comparison_table2)
```

```{r visualize_weighted_slope_models, include = TRUE}
bayesian_r2.1 <- bayes_R2(no_weighted_offset2)
b2_crash <- bayesian_r2.1[1, "Estimate"]

# Step 1: Create a finer grid of mean_temp for smoother prediction lines (without 'site')
prediction_grid2 <- tibble(mean_temp = seq(min(model_data_recover$mean_temp), max(model_data_recover$mean_temp), length.out = 100))

# Step 2: Generate predictions using the fitted values from the linear model
predicted_slope_values1 <- fitted(no_weighted_offset2, newdata = prediction_grid2, summary = TRUE)[, "Estimate"]

# Step 3: Add the predicted values to the dataset
prediction_grid2 <- prediction_grid2 %>%
  mutate(predicted_crash = predicted_slope_values1)

# Step 4: Plot with a straight line for the linear model
ggplot(model_data_recover, aes(x = mean_temp, y = slope_weighted)) +
  geom_point(aes(size = last_count), alpha = 0.5) +  # Plot observed data
  geom_line(data = prediction_grid2, aes(x=mean_temp, y = predicted_crash), color = "black", linewidth = 1) +
  scale_size_continuous(name = "Last\nSurvey\nCount") + 
  labs(title = "Mines with colder mean temperatures have lower crash rates for Myotis Bats",
       x = "Mean Temperature (C)",
       y = "weighted slope (slope * sqrt(last_count))") +
annotate("text", x = max(model_data_recover$mean_temp) - 0.5, 
         y = max(model_data_recover$crash) - 0.05, 
         label = paste("Bayesian R² =", round(b2_crash, 4)), 
         hjust = 3.25, vjust = 3.65, size = 4, color = "black") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12),        # Title font size
    axis.title.x = element_text(size = 10),      # X-axis title font size
    axis.title.y = element_text(size = 10),      # Y-axis title font size
    axis.text.x = element_text(size = 8),        # X-axis text font size
    axis.text.y = element_text(size = 8),        # Y-axis text font size
    legend.title = element_text(size = 10),      # Legend title font size
    legend.text = element_text(size = 8),
    panel.grid = element_blank()        # Legend text font size
  )
```